### 2025-05-11 thesis/thesis.tex
Write a section on ICL

In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
research focuses on the recent advancements when it comes to language models and how they can be harnessed for
effective NL2SQL systems \citep{LLM-Sql, T2SQL-LLM-Bench, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.

@@

% https://raw.githubusercontent.com/HKUSTDial/NL2SQL_Handbook/refs/heads/main/assets/river.svg
% https://arxiv.org/abs/2109.05093

@@

% CodeS
% GPT-X based
% 
% Prompt engineering and fine tuning

%\subsection{Optimization for LLM-based NL2SQL}
% PICARD, DIN-SQL, MAGIC & CHASE

@@

\nocite{*}

>>>>

In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
research focuses on the recent advancements when it comes to language models and how they can be harnessed for
effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.

@@

The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
less of a central problem to solve \citep{DINSQL}.

@@

The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.

\subsubsection{In-Context Learning}

In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.

The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
/ prior usage of the system can be dynamically utilized, rather than requiring retraining.

Example selection strategies showed to have a significant impact on ICL performance. \cite{DAIL-SQL} evaluated
various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.

The comparison between zero-shot and few-shot performance reveals the significant impact of supplying in-context examples
to models. While LLMs demonstrate impressive zero-shot capabilities, with models like GPT-4 achieving 72.3\% execution
accuracy on \textsc{Spider} without any examples \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning shows to substantially
improve execution accuracy. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's execution accuracy to 80.2\%,
representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}. This improvement
in execution accuracy is particularly noticable for complex queries which involve multiple joins or nested subqueries,
where zero-shot approaches often fail generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the
leap in exact match ratio measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to
71.9\% with five-shot. The results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution
accuracy of $k$-shot approaches.

The effectiveness of ICL approaches has established them as the standard technique applied in LLM-based NL2SQL systems.
They provide a robust foundation which can be extended with advanced techniques such as problem-decomposition, self-correction
and chain-of-thoughts. ICL alone is still subject to frequent limitations observed with NL2SQL systems like ambiguoity handling,
schema linking and dealing with complex queries. This motivated further research of LLM-based NL2SQL systems.

@@
